---
title: "The Impacts of Inflation and Interest Rates on the Stock Market"
subtitle: "Report"
format: html
editor: visual
execute:
  echo: true
---

```{r message = FALSE}
#|: loading-packages

library(tidyverse)
library(tidymodels)
library(scales)
```

# **Introduction**

We want to explore the relationship between interest and inflation rates and the performance of the stock market. In doing so, we hope to determine economic characteristics in which it's beneficial to invest in the stock market, as well as if/when certain companies or sectors perform better in different economic climates. Therefore, we are interested in answering the following questions:

-   How have these rates (inflation and interest) historically impacted the returns of the stock market?

-   How closely has the current performance of the market (the past two years) followed these historical trends?

-   And finally, does the current high-rate economic climate disproportionately affect some companies/sectors more than others? If so, why?

We believe the overall stock market would be hurt by high inflation and corresponding high interest rates, both because consumers' discretionary incomes drop and because companies are disincentivized from borrowing, due to the higher costs associated with taking out a loan. We expect to see this relationship in both our historical analysis and our current-day analysis. Conversely, in times of lower rates, we expect market returns to be higher, on average. Though we expect the stock market as a whole to suffer from high rates, we believe certain sectors of the economy could serve as silver linings for investors' pockets; these industries include: financials, health care, utilities, and consumer staples. We believe these sectors could perform better than the aggregate market because they sell essential goods that consumers must purchase no matter the state of the economy (health care, utilities, and consumer staples). Also, we believe the financials sector will do well because banks are able to charge higher interest rates on their loans (and thus earn greater profits) when interest rates are high.

Note: We are using the S&P 500 as a model of the performance of the entire market, and we are using the federal-funds rate, the interest rate at which banks loan to other banks, to model the general trend of all interest rates. We believed these are logical assumptions, as the S&P 500 includes the 500 largest US-based companies (and thus covers a large portion of the total market) and the federal-funds rate generally serves as a benchmark for all other interest rates (because it is indirectly influenced by the Federal Reserve's policies). Also, all of our observations feature the performance of the market (S&P 500) or an individual company in the S&P 500 on a specific day/month/year.

\newpage

# Part 1: Historical Analysis

Sources:

-   Federal Funds Rate and Inflation Rate: Macrotrends (<https://www.macrotrends.net/2015/fed-funds-rate-historical-chart>)

-   S&P 500 Annual Returns: Slickcharts (https://www.slickcharts.com/sp500/returns)

-   Monthly Statistics (scraped): Official Data (https://www.officialdata.org/us/stocks/s-p-500/1900).

All of these sites collected these data by retrieving the data published publicly on stock exchanges, except for the federal-funds rate--this data was collected from the World Bank. In addition, each of these sites updates their data monthly/yearly (although they could update it daily, as they are collecting live data). Each observation represents a month in a year (between 1961 and 2020) and includes the following variables:

| Variable \| Definition \|

\|---------------------\|------------------------------------------------------\|

year \| year in which observation was recorded \| \| month \| month in which observation was recorded \| \| monthly_return \| total stock market's monthly return \| \| year_to_date_return \| S&P 500's yearly return \| \| mean_fed_funds_rate \| mean annual interest rate set by the Federal Reserve \| \| inflation rate \| mean annual inflation rate \| \| decade \| represents decade in which observation was made \|

```{r message = FALSE}
#|: loading-historical-data

historical_data <- read_csv("data/historical_data.csv")
glimpse(historical_data)
```

```{r}
#|: inflation-vs-interest-plot
# Completed with assistance from: https://community.rstudio.com/t/adding-manual-legend-to-ggplot2/41651/5

colors <- c("Inflation" = "blue", "Interest" = "red")

ggplot(historical_data) +
  geom_line(aes(x = year, y = mean_fed_funds_rate, color = "Interest")) +
  geom_line(aes(x = year, y = inflation_rate, color = "Inflation")) +
  theme_minimal() +
  labs(
    title = "Interest Rates vs. Inflation Rates",
    subtitle = "1961 - 2020",
    x = "Year",
    y = "Average Annual Rate (%)",
    color = "Type"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    legend.title = element_text(hjust = 0.5)
  ) +
  scale_color_manual(values = colors) +
  scale_color_viridis_d()
```

We first isolated interest and inflation rates to determine the relationship between them. To do so, we utilized a line graph to see their trends across time--more specifically, from 1961 to 2020. As you can see in the graph above, these rates are very closely related (especially prior to 2000), with interest and inflation peaking and dipping at similar times. This trend makes sense, as the Federal Reserve tends to raise the federal-funds (interest) rate when inflation rises, in an attempt to slow the economy down. (This is why the Federal Reserve is raising interest rates now--more on that later.)

```{r}
#|: interest-inflation-and-market-returns-graph

colors_redux <- c("Inflation" = "blue", "Interest" = "red", "Market Returns" = "green")

historical_data |>
  ggplot() +
  geom_line(aes(x = year, y = mean_fed_funds_rate, color = "Interest")) +
  geom_line(aes(x = year, y = inflation_rate, color = "Inflation")) +
  geom_line(aes(x = year, y = year_to_date_return, color = "Market Returns")) +
  theme_minimal() +
  labs(
    title = "Impacts of Interest Rates and Inflation Rates on Market Returns",
    subtitle = "1961 - 2020",
    x = "Year",
    y = "Average Annual Rate (%)",
    color = "Type"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    legend.title = element_text(hjust = 0.5)
  ) +
  scale_color_manual(values = colors_redux) +
  scale_color_viridis_d()
```

Then, we analyzed the relationship between rates and average yearly returns of the stock market for the same period using another line graph. As depicted in the graph, market returns are extremely volatile and (at least according to this graph) doesn't seem to depend on either interest or inflation, contrary to our hypothesis.

```{r}
#|: best-ten-years-of-returns

historical_data_best <- historical_data |> 
  distinct(year, mean_fed_funds_rate, inflation_rate, year_to_date_return) |>
  arrange(desc(year_to_date_return)) |>
  slice_head(n = 10) |> 
  relocate(year, year_to_date_return)

historical_data_best
```

```{r}
#|: worst-ten-years-of-returns

historical_data_worst <- historical_data |> 
  distinct(year, mean_fed_funds_rate, inflation_rate, year_to_date_return) |>
  arrange(year_to_date_return) |>
  slice_head(n = 10) |> 
  relocate(year, year_to_date_return)

historical_data_worst
```

Next, we determined the best and worst 10 years of performance for the stock market by sorting by year_to_date_return, in order to examine the years individually rather than as a trend across time. Many of the best years feature at least one relatively low rate (\<5%), with many having both low interest and inflation rates. Noticeable outliers include 1980, whose rates were both over 13%, and 1975, whose interest rate was almost 6% and inflation 9%. This trend of high returns correlating with lower rates does support our hypothesis. On the other hand, we expected many of the worst years for market returns to have high rates; surprisingly, this was not the case for about half of the worst years. Though some years did support our hypothesis, including 1974, 1973, 1969, and 1977, the others featured surprisingly low rates. We suspect that these unanticipated poor performance of the stock market during times of low rates could possibly be explained by recessions occurring during these years; in recessionary gaps, inflation tends to be lower because the economy has slowed significantly, and interest rates tend to be low as well in an attempt to re-stimulate the economy by encouraging borrowing.

```{r}
#|: creating-df-of-averages-of-returns-and-rates-by-decade
# excluded 2020 because it's not a whole decade

historical_data_decade <- historical_data |>
  filter(decade != "2020") |>
  group_by(decade) |>
  summarize(
    mean_return = mean(year_to_date_return),
    mean_interest = mean(mean_fed_funds_rate),
    mean_inflation = mean(inflation_rate)
  ) |>
  arrange(desc(mean_return))
```

```{r}
#|: visualization-of-rates-by-decade

historical_data_decade |>
  pivot_longer( 
    !decade, 
    names_to = "type",
    values_to = "rate"
  ) |>
  mutate(
    type = as.factor(type),
    type = fct_rev(type)
  ) |>
  ggplot(aes(x = type, y = rate, fill = type)) +
  geom_col() +
  facet_wrap(~decade, ncol = 3) +
  labs(
    title = "Mean Returns and Rates by Decade",
    subtitle = "1960s - 2010s",
    y = "Mean Rate",
    fill = "Type"
  ) +
  scale_fill_viridis_d() + 
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    legend.title = element_text(hjust = 0.5),
    axis.title.x = element_blank(),
    axis.text.x = element_blank()
  )
```

For the last part of our historical analysis, we wanted to examine the averages of returns, interest, and inflation by decade. To do so, we created a bar chart for each decade (using geom_col() and facet_wrap()). This graph supports what we concluded after examining the best and worst 10 years of returns: higher returns seem to be common in times of low rates, but it's difficult to determine a relationship between returns and rates when returns are lower. For example, the '60s, '90s, and 2010s all featured high returns (\~10+%) and rates at or below 5%; an outlier decade is the 2000s, whose average return was an abysmal \~1% despite low rates. On the other hand, in decades when rates were high (the '70s and '80s), returns were much more volatile--decent in the '70s (7.5%) yet great in the '80s (15+%).

Conclusion: Our initial hypothesis was partially correct. It seems that high returns in the stock market generally correlate with lower interest and inflation rates, but years with lower returns can't be explained as neatly, at least with the data we're using. A potential confounding variable could be whether our economy was experiencing a recession during that year, as recessions sometimes create environments where returns in the stock market are poor even in times of low rates.

\newpage

# Part 2: Current-day Analysis

Source:

-   Investing.com (https://www.investing.com/indices/us-spx-500-historical-data)

Investing.com gets their data from financial-data providers, including stock exchanges and/or directly from market makers (who set stock prices to match supply and demand). The data is collected in real time by monitoring transactions (selling, buying, etc.) executed in the stock market, which correspond to increases/decreases in prices. Each observation represents a different day (between 1 October 2020 and 28 October 2022) and includes the following variables:

| Variable \| Definition \|

\|-------------------------\|-------------------------------------------------------------------\|

| date \| day on which observation was recorded \|
| price \| closing daily price for the S&P 500 \|
| open \| opening daily price for the S&P 500 \|
| daily_percentage_change \| daily percentage difference between opening and closing prices \|
| daily_diff \| daily numerical difference between opening and closing prices \|
| gain \| TRUE if daily_diff \> 0 (positive return for the day), else FALSE \|

Contextualization:

```{r message = FALSE}
#|: each-day-closing-price

sp500_curr <- read_csv("data/sp500_curr_updated.csv")
glimpse(sp500_curr)

sp500_curr

sp500_curr|>
#  mutate(Date = paste(substr(Date, 9, 10), substr(Date, 1, 2),  sep = "/"))|>
  ggplot(aes(x = Date, y = Price)) +
  geom_point()



```

```{r}
#| label: gains-and-losses-per-month
sp500_curr|>
  mutate(Date = paste(substr(Date, 9, 10), substr(Date, 1, 2),  sep = "/"))|>
  group_by(Date)|>
  count(gain) |>
  ggplot(aes(x = Date, y = n, fill = gain)) +
  geom_col(position = "fill") +
  labs(title = "Number of Days with Gains and Losses by Month", y = "# days", x = "Month (year/month)") +
  theme_minimal() +
  scale_color_viridis_d()

```

```{r}
#| l label: last-first-day-price-difference
sp500_curr

  
```

Conclusion:

\newpage

# Part 3: Company/Sector Analysis

Sources:

-   Name, Symbol, and Sector: DATA HUB (https://datahub.io/core/s-and-p-500-companies-financials#resource-constituents)

-   Weights (scraped): Slickcharts (<https://www.slickcharts.com/sp500>)

-   Prices and Returns (scraped): Slickcharts (<https://www.slickcharts.com/sp500/performance>)

DATA HUB collected their data from S&P Dow Jones Indices (which, in turn, is collected daily from stock exchanges), and Slickcharts their data directly from the stock exchange. Each observation represents a different company and includes the following variables:

| Variable        | Definition                                                                    |
|------------------|------------------------------------------------------|
| name            | name of company in S&P 500                                                    |
| symbol          | ticker symbol used to identify companies on stock exchange                    |
| weight          | percentage weighting of company in S&P 500                                    |
| original_price  | price of company's stock at beginning of 2022                                 |
| current_price   | price of company's stock as of October 2022                                   |
| ytd_return_num  | company's year-to-date return percentage                                      |
| sector          | company's industry                                                            |
| return category | positive if ytd_return_num \> 0 (positive return for the year), else negative |

```{r message = FALSE}
#|: loading-current-day-company-and-sector-data

sp500_curr_company_and_sector <- read_csv("data/sp500_by_company_and_sector.csv")
glimpse(sp500_curr_company_and_sector)
```

-   How many stocks rose or fell?

```{r}
sp500_by_company_and_sector <- read_csv("data/sp500_by_company_and_sector.csv")
```

```{r}
#|: overall-stocks_performance-barchart
sp500_by_company_and_sector |>
  ggplot(aes(x = return_category)) +
  geom_bar() +
  labs(
    title = "Rising and Decreasing Stock Prices",
    x = "Change in Price (%)",
    y = "Number of Companies",
    fill = "cc"
       )
```

-   How many of stocks rose or decreased in each sector?

```{r}
#|: stocks-by_sector_barchart
sp500_by_company_and_sector |>
  ggplot(aes(x = return_category)) +
  geom_bar() +
  facet_wrap(~sector) +
  labs(
    title = "Rising and Decreasing Stock Prices",
    x = "Change in Price (%)",
    y = "Number of Companies",
    fill = "cc"
       )
```

-   What is the variation in percentage returns for each sector?

```{r}
#|: sectors-performance-barchart
sp500_by_company_and_sector |>
  #group_by(sector) |>
  #summarize(sector_perf = mean(ytd_return_num, na.rm = TRUE)) |>
  #mutate(sector_ret = case_when(
  #     sector_perf > 0 ~ "Increased",
  #     sector_perf <= 0 ~ "Decreased")) |>
  ggplot(aes(y = sector, x = ytd_return_num)) +
  geom_boxplot() +
  labs(
  title = "Rising and Decreasing Stock Prices by Sector",
     x = "% Returns",
     y = "Sector",
     fill = "cc"
        )
```

-   What are the top performing sectors?

```{r}
sp500_by_company_and_sector |>
  group_by(sector) |>
  summarize(sector_perf = mean(ytd_return_num, na.rm = TRUE)) |>
  arrange(desc(sector_perf))
```

-   What are the top 10 companies with the best returns and what sector are they in?

```{r}
#|: 10_best_returns
sp500_by_company_and_sector |>
  arrange(desc(ytd_return_num)) |>
  slice_head(n = 10) |>
  select(name, symbol, ytd_return_num, sector)
```

-   What are the top 10 companies with the worst returns and what sector are they in?

```{r}
#|: 10_worst_returns
sp500_by_company_and_sector |>
  arrange(ytd_return_num) |>
  slice_head(n = 10) |>
  select(name, symbol, ytd_return_num, sector)
```

Conclusion:
